---
title: "Project 3 Data Science NanoDegree "
author: "Fang Lu"
date: "Monday, May 25, 2015"
output: html_document

---

 
**Note:** This is currently a placeholder for my Project 3, I will be completing this project within 48 hours of submission (end of day May 27, 2015), Please keep me with my current Decemeber cohort. I have gotten somewhat behind on project 3, but should have no issues completing the rest of the projects with my cohort. Thank you for your consideration.



###1. Injestion and Familiarization of the Data

This is an exploration of a tidy data set provided by Udacity Data Science Nanodegree for Project 3.

```{r}
summary(cars)
```

You can also embed plots, for example:

```{r, echo=FALSE}
plot(cars)
```

###2. Free Exploration

```{r}
summary(cars)
```


###3. Further Exploration

```{r}
summary(cars)
```


###4: Final Plots and Summary

```{r}
plot(cars)
```

###5: Reflection

Relections are what we see in mirrors.

```{r}
summary(cars)
```


