---
title: "Project 3 Data Science NanoDegree "
author: "Fang Lu"
date: "11/16/2015"
output: html_document

---
```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE, fig.width = 12, fig.height = 8)
```


###1. Injestion and Familiarization of the Data

This is an exploration of the tidy data set **wineQualityReds.csv** provided by Udacity Data Analyst Nanodegree for Project 3. This data set was chosen for the brevity of the observations in consideration for execution time of certain plots. Guiding question: "Which Variables affect Red Wine Quality?"


```{r, include=FALSE}
#This block has been set so that the library loading does not show up in the
#knit html
#Make sure that one is working in the correct working directory
#getwd()
#setwd('C:/Fang Lu/Udacity/Data Science/EDA_Course_Materials/EDA_Course_Materials/Project3')
#Read in Wine Quality Reds as wqr and run all necessary libraries
library(ggplot2)
library(GGally)
library(gridExtra)
library(reshape)
library(scales)
library(memisc)
library(RColorBrewer)
library(stats)
library(tidyr)
library(dplyr)
wqr <- read.csv('wineQualityReds.csv')

```

*Initial Explorations:*

```{r Summaries}
names(wqr)
str(wqr)
summary(wqr)
```

Some parameter names may be a bit long to plot on ggpairs, perhaps a renaming of the parameters will be needed later. There are 1599 observations of red wine in the data-set (rather small considering other data-sets). Alcohol content of red wines range from 8.40% to 14.90% with most around 10%. The quality rating of reds in this data set are mostly between 5 and 6 with median at 6. pH of reds are stable around 3-4, 

###2. Free Exploration and Transformation

```{r Early Exploration}
#Added jitter and some transparency to the basic plot
ggplot(aes(x = quality, y = fixed.acidity), data = wqr) + 
  geom_point(alpha = 1/4, position = 'jitter')

```
------ 
It seems that fixed acidity has little to do with the quality of red wines, and that there are more quality 5 to 7 wines than there qualities of other types, this is also noted in the readme *wineQualityInfo.txt*. Perhaps it would be better to combine the bottom two levels and the top two levels.

```{r Histograms, message=FALSE}
#Early Histogram Generation to get an idea of the distribution of parameters
h1 <- ggplot(aes(x = fixed.acidity), data = wqr) + 
  geom_histogram(binwidth = .25)
h2 <- ggplot(aes(x = sqrt(volatile.acidity)), data = wqr) +
  geom_histogram(binwidth = .05)
grid.arrange(h1, h2, ncol = 2)

#Using Reshape2 to melt the data together to generate a histogram gridplot for 
#an overview
meltwqr <- melt(wqr, id = c('X'))
gridHist <- ggplot(aes(x=value), data=meltwqr) + 
  geom_histogram() + 
  facet_wrap(~variable, ncol = 4, scale='free_x', shrink=TRUE)
gridHist

```
------ 
The histogram grid took some research to create (at first attempted a function that did not work as intended). This provided a good overview of the distribution of the different chemical attributes, *residual.sugar* and *chlorides* and perhaps *sulphates* could possibly use a closer look at the X-scaling for they seem to be more long-tailed. An extra note, density seems to be rather normally distributed.

```{r Scatter Plots, message=FALSE}
#created a second transformation of the data using melt from reshape2 library to
#get a scatter plot of all the input variables vs quality
scattwqr <- melt(wqr, id = c('X','quality'))
gridScatt <- ggplot(aes(x=quality, y = value), data = scattwqr) + 
  geom_point(alpha = 1/4, position= "jitter", size=0.5) + 
  facet_wrap(~variable, ncol=4, scale='free_y')
gridScatt

```
------ 
From the scatter plots, *sulphates*, *alcohol* content seems to have a bit of correlation with quality where higher alcohol content seems to indicate higher quality, but the variance is still pretty high.

```{r boxplots, message=FALSE}
#Boxplots for graphical statistical summary of all the input variables vs
#quality
gridBox <- ggplot(aes(x=factor(quality), y = value), data = scattwqr) + 
  geom_boxplot() + facet_wrap(~variable, ncol=4, scale='free_y')
gridBox
```
------ 
The boxplots seem to reveal some interesting trends where the other plots showed relatively little. A re-scaling of the Y-axis for *residual.sugar*, *chlorides*, and *sulphates* might reveal a better view of the boxplots.

```{r Detailed Summaries, message=FALSE}

#with(wqr, cor.test(quality, citric.acid, method = 'pearson'))
#with(wqr, cor.test(quality, volatile.acidity, method = 'pearson'))
#with(wqr, cor.test(quality, fixed.acidity, method = 'pearson'))
#with(wqr, cor.test(quality, alcohol, method = 'pearson'))
#with(wqr, cor.test(quality, sulphates, method = 'pearson'))
#with(wqr, cor.test(quality, chlorides, method = 'pearson'))
#with(wqr, cor.test(density, alcohol, method = 'pearson'))
#with(wqr, cor.test(density, quality, method = 'pearson'))
with(wqr, cor.test(quality, fixed.acidity, method = 'pearson'))
#No need to sample since our dataset is relatively small
#create a new dataframe with condensed naming for ggpairs
newnames = c('X', "f.acid",'v.acid','c.acid','r.sugar', 'clrides', 
             'f.s.dioxide','total.s.d', 'density','pH','sulphates',
             'alcohol','quality')
wqrRen <- wqr
colnames(wqrRen) <- newnames
#ggpairs(wqrRen[3:13], params = c(shape = I('.'), outlier.shape = I('.')))

#run ggpairs for 10 of the variables, see newnames for the column names.
ggpairs(wqrRen[3:13], upper = list(params = c(size = 4)), 
        lower = list(continuous = "smooth"))

#Some sanity checks

with(wqr, cor.test(fixed.acidity, pH))

```

A sanity check with *fixed acidity* vs *pH* shows that as expected there is a reasonably high correlation: -.68. Looking at the correlation coefficients, it seems that *alcholol*, *volatile acidity*, *sulphates*, and *acid acid* have the four highest correlation with *quality*. *pH*, *residual sugar* and *free sulfur dioxide* are seemingly uncorrelated with *quality*.

**Citric Acid**  

```{r Single Plots}

summary(wqr$citric.acid)

ggplot(aes(x = citric.acid), data = wqr) + geom_histogram() + 
  scale_x_continuous(limits = c(0, 1.2), 
                     breaks = c(0, .15, .3, .45, .6, .75, .8, 1, 1.1))

ggplot(aes(x = citric.acid), data = wqr) + 
  geom_histogram(binwidth = .01,color = 'black', fill='blue') + 
  scale_x_continuous(limits = c(0, .85), breaks = seq(0,.85,.03))

wqr$cAcid.cut <- cut(wqr$citric.acid, breaks = c(.001,.15,.30,.50,1.2))
levels(wqr$cAcid.cut)

with(wqr, cor.test(quality, as.numeric(cAcid.cut), method = 'pearson'))

```

Created the categorical variable *cAcid cut* from citric.acid by looking at the histogram (which is relatively evenly distributed) and then maximizing the correlation. I feel that more categorical variables could help simply and visualize how each attribute may be attributing to the *quality* factor. Also, correlation increases with the new categorical variable.

**Free and Total Sulfur Dioxide**  

```{r}
#The readme gave hints as to the cutoffs to free total sulfur dioxide to a 
#categorical variable. Perhaps cut total sulfur dioxide as well might affect 
#taste since it's correlated with free sulfure dioxide
ggplot(aes(x=total.sulfur.dioxide), data = wqr) + 
  geom_histogram(binwidth = 1, color = 'black',fill='green') + 
  xlim(0,100)

#Chose the cutoff levels experimentally maximizing the correlation. Started with
#~50 and 3 levels moved 4 and tweaked the 50 cutoff to 59
wqr$total.sulf.dioxide.cut <- cut(wqr$total.sulfur.dioxide, 
                                  breaks = c(0,59,109,300))

levels(wqr$total.sulf.dioxide.cut)

with(wqr, cor.test(quality, as.numeric(total.sulf.dioxide.cut), 
                   method = 'pearson'))

ggplot(aes(x=free.sulfur.dioxide), data = wqr) + 
  geom_histogram(binwidth = 1, color = 'black',fill='green')+xlim(0,100)

summary(subset(wqr, wqr$free.sulfur.dioxide>=50))

wqr$free.cut <- cut(wqr$free.sulfur.dioxide, breaks = c(0,10,20,30,40,50,100))

with(wqr, cor.test(quality, as.numeric(free.cut), method = 'pearson'))

ggplot(aes(y=quality, x=free.cut),data=wqr) + 
  geom_point(size=3,position='jitter',alpha=.5)
```

The readme hinted that there might be adverse tastes when free sulfur dioxide values becomes greater than 50ppm. There was an increase in correlation from the numerical version of *'free.sulfur.dioxide'* parameter to the categorical one, perhaps it will be useful in gaining insight to the quality parameter? I also turned *'total.sulfure.dioxide'* variable into a categorical variable (saw an increase in correlation with *'quality'*)

**Chlorides and Density**  

```{r}
#Chlorides and Density a closer look
ggplot(aes(x=chlorides), data = wqr)+geom_histogram(binwidth = .001)+
  scale_x_continuous(limits = c(0, .15), breaks = seq(0,.15,0.01))

wqr$chloride.cut<- cut(wqr$chlorides, breaks = 
                         c(0, quantile(wqr$chlorides, 0.25), 
                           quantile(wqr$chlorides, 0.5), 
                           quantile(wqr$chlorides, 0.75), .7))

levels(wqr$chloride.cut)
with(wqr, cor.test(quality, as.numeric(chloride.cut), method = 'pearson'))

```

Both Density and chlorides have very normal distributions, and have reasonable correlations with quality. After resizing the x-axis, it seems that chlorides is a very normally distributed variable. Turning chloride into a categorical variable increased it's correlation with quality from -0.12 to -0.17. 

```{r}
#Density very normally distributed

wqr$density.cut<- cut(wqr$density, breaks = c(0, quantile(wqr$density, 0.33), 
                                              quantile(wqr$density, 0.66),1.1))
levels(wqr$density.cut)

with(wqr, cor.test(quality, as.numeric(density.cut), method = 'pearson'))

```

Both new categorical variables show an increased correlation with quality. Does this mean if a regression model is made, the categorical inputs will yield a better model?


**Rabbit Hole Idea**  

```{r Down the Rabbit Hole only to find NOTHING, echo=TRUE}

wqr$dcut<- cut(wqr$density,
  breaks = c(0, quantile(wqr$density, 1/12),quantile(wqr$density, 2/12),
             quantile(wqr$density, 3/12),quantile(wqr$density, 4/12),
             quantile(wqr$density, 5/12),quantile(wqr$density, 6/12),
             quantile(wqr$density, 7/12),quantile(wqr$density, 8/12),
             quantile(wqr$density, 9/12),quantile(wqr$density, 10/12),
             quantile(wqr$density, 11/12),quantile(wqr$density, 12/12)))

levels(wqr$dcut)

levels(wqr$dcut)<-c("1","2","3","4","5","6","6","5","4","3","2","1")

with(wqr, cor.test(quality, as.numeric(dcut), method = 'pearson'))
```

There was a thought that evenly distributed variables like density and pH despite the low correlation might have more to say. The idea is simple, "high density, or pH or low density or pH" will possibly produce 'bad' quality wines (extremes tend toward the negative outcome in other words). As wine makers try to create 'good' wines, the independent variables should have a tendency toward the middle of the distribution. By cutting the data into 12 equal parts and then combining the ends (high and low pairs) together, the new categorical variable might reveal better correlation with the *'quality'* factor. A slight drop in correlation was found :(.  

```{r Down Down}
ggplot(aes(x=density, y=quality, color=dcut), data=wqr) + 
  geom_point(alpha = 1/2, position="jitter", size=3)


ggplot(aes(x=density, y=quality), data=wqr) + 
  geom_point(alpha = 1/2, position="jitter", size=3)

ggplot(aes(x=dcut, y=quality), data=wqr) + 
  geom_point(alpha = 1/2, position="jitter", size=3)
```

Something I was looking for in the plots were perhaps a higher number of low quality rankings associated category 1 of the dcut (which is the combination of the highest and lowest density after cutting the samples into 12 evenly populated segments). Plots do not reveal anything significant, although very colorful.  Perhaps pH will reveal something.  


```{r Rabbit Hole Continued, echo=TRUE}

wqr$pHcut<- cut(wqr$pH, 
                breaks = c(0, quantile(wqr$pH, 1/12),quantile(wqr$pH, 2/12),
                           quantile(wqr$pH, 3/12),quantile(wqr$pH, 4/12),
                           quantile(wqr$pH, 5/12),quantile(wqr$pH, 6/12),
                           quantile(wqr$pH, 7/12),quantile(wqr$pH, 8/12),
                           quantile(wqr$pH, 9/12),quantile(wqr$pH, 10/12),
                           quantile(wqr$pH, 11/12),quantile(wqr$pH, 12/12)))
levels(wqr$pHcut)

levels(wqr$pHcut)<-c("1","2","3","4","5","6","6","5","4","3","2","1")

with(wqr, cor.test(quality, as.numeric(pHcut), method = 'pearson'))

with(wqr, table(factor(quality), pHcut))

with(subset(wqr, as.numeric(wqr$dcut)>=5), table(factor(quality), pHcut))

```

Not looking good with pH, maybe a combination of pH and density will reveal an unexpected correlation. where the comination of pH and density may reveal some clustering of higher or lower quality  

```{r}
ggplot(aes(x = pH, y = quality, color = pHcut), data = wqr) + 
  geom_point(alpha = 1/2, position = "jitter", size = 3)

ggplot(aes(x = pHcut, y = quality, color = dcut), data = wqr) + 
  geom_point(alpha = 1/2, position = 'jitter', size = 3) + 
  ggtitle("Despair Never Looked so... Colorful, Oh Maybe that's the Rabbit") +
  scale_color_brewer(type = 'seq', palette = 'Blues',
    guide = guide_legend(title = 'Density (dcut)', reverse = F,
    override.aes = list(alpha = 1, size = 1)))


ggplot(aes(color = pHcut, y = quality, x = dcut), data = wqr) + 
  geom_point(alpha = 1/2, position = 'jitter', size = 3) +
  ggtitle("Despair Never Looked so... Colorful PART 2") +
  scale_color_brewer(type = 'seq', palette = 'Blues',
    guide = guide_legend(title = 'pH', reverse = F,
    override.aes = list(alpha = 1, size = 1)))

```

pH investigation dead-end, though I still haven't given up that there might be something in the normally distributed data. Maybe Chlorides and residual sugar have some story to tell.

```{r What is Left of my Soul}

ggplot(aes(x=factor(quality), y = residual.sugar), data = wqr) + 
  geom_boxplot()+ylim(0,4)

ggplot(aes(x=factor(quality), y = chlorides), data = wqr) + 
  geom_boxplot()+ylim(0,0.2)


printVariance <- function(colname,data){
  for (i in 3:8){
    print(with(subset(get(data), get(data)$quality == i), var(get(colname))))
  }
}

printVariance("residual.sugar",'wqr')
printVariance("chlorides",'wqr')
printVariance("pH",'wqr')
printVariance("density",'wqr')

summary(wqr$chlorides)
```

Can people actually taste .1 g/liter of chloride/salt? Note: our chloride and residual sugar measurements are in g/dm^3 which is the same as g/liter. This might be a futile exercise. Looked at variances of several variables to see if there is a trend of increasing or decreasing variance. This could indicate that the tails of the distribution might contain correlation to one particular quality level once again the 'Too much or too little yields bad quality' theory.  

```{r Please Chlorides Please..., echo=TRUE}
wqr$chlcut<- cut(wqr$chlorides, 
  breaks = c(0, quantile(wqr$chlorides, 1/12),quantile(wqr$chlorides, 2/12),
             quantile(wqr$chlorides, 3/12),quantile(wqr$chlorides, 4/12),
             quantile(wqr$chlorides, 5/12),quantile(wqr$chlorides, 6/12),
             quantile(wqr$chlorides, 7/12),quantile(wqr$chlorides, 8/12),
             quantile(wqr$chlorides, 9/12),quantile(wqr$chlorides, 10/12),
             quantile(wqr$chlorides, 11/12),quantile(wqr$chlorides, 12/12)))
levels(wqr$chlcut)

levels(wqr$chlcut)<-c("1","2","3","4","5","6","6","5","4","3","2","1")

summary(wqr$chlorides, 10)

ggplot(aes(x=chlorides, y=quality, color=chlcut), data=wqr) + 
  geom_point(alpha = 1/2, position="jitter", size=3)

ggplot(aes( y=quality, x=chlcut), data=wqr) + 
  geom_point(alpha = 1/2, position='jitter', size=3) +
  ggtitle("Despair Never Looked so... Colorful Why am I still doing this") +
  scale_color_brewer(type = 'seq', palette = 'Blues',
    guide = guide_legend(title = 'pH', reverse = F,
    override.aes = list(alpha = 1, size = 1)))
with(wqr, table(factor(quality), chlcut))

```

The wikipedia article on "Taste" revealed that the average human detection threshold for sucrose is 10 millimoles per liter which actually translates to .01g/liter. Further searching has some studies showing that salt detection threshold is around .5 mols/liter which translates to .5 grams/liter. Most likely the difference of salts contained in these wines are not at levels that are able to be readily distinguished by humans with a range of (.012-.611 g/liter). The differences in sugar should be able to be detected however. But once again, the 'folding' of the distribution has not revealed any new correlations.

###3. Further Exploration some Multivariate

**Alcohol**
```{r Alcohol}

#Visualization of the relationship between Alchohol content and Quality

ggplot(aes(x=quality, y=alcohol), data = wqr) + 
  geom_point(alpha=1/4, position='jitter', size = 3) + 
  stat_summary(fun.y = median, geom = 'point', 
               shape = 7, size = 10, color = 'red') 

```

Alcohol vs quality (a little backwards in the axis but it seems more natural to place quality on the x for visualization purposes). Added the median to see the the possible linear relationship with quality.

```{r Less is Better Maybe}
wqr$qual.cut <- cut(wqr$quality, breaks = c(2,4,5,6,7,8))
levels(wqr$qual.cut)

ggplot(aes(x=qual.cut, y=alcohol, color = cAcid.cut), data = wqr) + 
  scale_color_brewer(type = 'div',
    guide = guide_legend(title = 'Citric Acid', reverse = T,
    override.aes = list(alpha = 1, size = 1))) +
  geom_point(alpha=1/2, position='jitter', size = 3) + 
  stat_summary(fun.y = median, geom = 'point', 
               shape = 7, size = 10, color = 'red')

with(wqr, cor.test(as.numeric(qual.cut), alcohol, method = 'pearson'))
```

Created a new categorical variable out of *'quality'* called *'qual.cut'* by merging the bottom 2 categories in *'quality'*. Hope to make trends a bit clearer and provide a cleaner view of the bins.

**Volatile Acidity**
 
```{r Boom}
#Better Visualize the relationship between Volatile Acidity and Quality
ggplot(aes(x=factor(quality), y=volatile.acidity), data = wqr) +
  geom_point(alpha=1/4, position='jitter', size = 2)+
  geom_boxplot(alpha = .25)+
  stat_summary(fun.y = median, geom = 'point', 
               shape = 7, size = 6, color = 'red')

#Facet on Citric Acid
ggplot(aes(x=qual.cut, y=volatile.acidity, color = cAcid.cut), data = wqr) + 
  scale_color_brewer(type = 'div',
    guide = guide_legend(title = 'Citric Acid', reverse = T,
    override.aes = list(alpha = 1, size = 1))) +
  geom_point(alpha=1/2, position='jitter', size = 3) + 
  stat_summary(fun.y = median, geom = 'point', shape = 7, 
               size = 6, color = 'red') + ylim(0, 1.2)

#Volatile Acidity colored by Total Sulfure Dioxide categorical Variable
ggplot(aes(x=qual.cut, y=volatile.acidity, color = total.sulf.dioxide.cut), 
       data = wqr) + 
  scale_color_brewer(type = 'seq', palette = 'YlGnBu',
    guide = guide_legend(title = 'Total Sulfur Dioxide', reverse = F,
    override.aes = list(alpha = 1, size = 1))) +
  geom_point(alpha=.75, position='jitter', size = 3) + 
  stat_summary(fun.y = median, geom = 'point', 
               shape = 7, size = 6, color = 'red') + ylim(0, 1.2)


```
 
The negative correlation between Volatile Acidity and Citric Acid can be seen the above histogram. The higher the Volatile Acidity, the lower the Citric Acid metric. There seems to be a reasonably strong correlation shown between Total Sulfur Dioxide and quality.

**Sulphates** 

```{r High Correlation}
ggplot(aes(x=qual.cut, y=sulphates), data = wqr) + 
  geom_point(alpha=1/4, position='jitter', size = 3) +
  stat_summary(fun.y = median, geom = 'point', shape = 7, 
               size = 6, color = 'red')+coord_cartesian(ylim=c(.25,1))


ggplot(aes(x=qual.cut, y = sulphates), data =wqr) + 
  geom_boxplot() + coord_cartesian(ylim=c(.25,1)) + 
  stat_summary(fun.y = mean, geom = 'point', shape = 9, size = 6, color = 'red')

```

Sulphates seem to be linearly correlated with quality as shown by the ggpairs correlation matrix this is a good variable add to a model.


```{r Extra Sweet and Sour}
#Fixed Acidity vs Sugar colored by quality
ggplot(aes(x=log(residual.sugar), y = fixed.acidity, color=factor(quality)), 
       data = wqr)+ geom_point(size = 3, alpha = .5)+
  scale_color_brewer(type = 'seq', palette = 'Blues',
    guide = guide_legend(title = 'Quality', reverse = F,
    override.aes = list(alpha = 1, size = 3)))

```

Feeling somewhat unsatisfied with the current findings and the lack of elements to account for the variance in the quality ratings (although quality ratings are subjective and discrete), I started reading about how wine is rated. There are many subjective measures in rating wines, but the one that makes the most sense is Appearance (Visual), Aroma(smell), *Taste*, and Aftertaste (finish). Perhaps it's best to see how good of a model we can obtain from the current inputs. 

**Modeling the Data**
```{r lm model}
m1 <- lm(I(quality) ~ alcohol, data = wqr)
m2 <- update(m1, ~ . + sulphates)
m3 <- update(m2, ~ . + volatile.acidity)
m4 <- update(m3, ~ . + cAcid.cut)
m5 <- update(m4, ~ . + total.sulf.dioxide.cut)
m6 <- update(m5, ~ . + chloride.cut)
m7 <- update(m6, ~ . + density.cut)
m8 <- update(m7, ~ . + pH)
m9 <- update(m8, ~ . + fixed.acidity)
m10 <- update(m9, ~ . + residual.sugar)
m11 <- update(m10, ~ . + free.sulfur.dioxide)

mtable(m1, m2, m3, m4, m5, m6, m7, m8,m9,m10,m11)

```

I decided to run a linear regression test to see how much of the variance in quality I can capture. The result is about 36% of the variance with all independent variables modeled. The same model was run with the original non-categorical variables, the resulting model was about 1-2% worse. There are several considerations:

1.  *'quality'* is essentially a categorical variable and should not be treated as a continuous one even though it looks like one. 
2. A linear model might not be the best model, perhaps a non-linear approach could yield better results
3. Perhaps the data does not capture enough of the variance as there are other subjective measures that *wine experts* take into account when rating wines such has Appearance and Aroma

```{r glm Model and Binomial Distribution Model for Quality}
#Take into account that quality is a categorical variable and perform a 
#generalized linear regression with probit
m1 <- glm(factor(quality) ~ alcohol, family = binomial(link = "probit"), 
          data = wqr)
m2 <- update(m1, ~ . + sulphates)
m3 <- update(m2, ~ . + volatile.acidity)
m4 <- update(m3, ~ . + cAcid.cut)
m5 <- update(m4, ~ . + total.sulf.dioxide.cut)
m6 <- update(m5, ~ . + chloride.cut)
m7 <- update(m6, ~ . + density.cut)
m8 <- update(m7, ~ . + pH)
m9 <- update(m8, ~ . + fixed.acidity)
m10 <- update(m9, ~ . + residual.sugar)
m11 <- update(m10, ~ . + free.sulfur.dioxide)

mtable(m1, m2, m3, m4, m5, m6, m7, m8,m9,m10,m11)

```

The generalized linear model treating the dependent variable *'quality'* as a categorical variable yield a better model, at least in the R-squared value sense yielding a McFadden R-sq. of .481 which seems to indicate that the model is a good fit for the data (non-categorical stands at R-sq of .430). I still wonder if a non-linear approach might be able to model the data more effectively.

**Sulfur Dioxide and Sulphate Exploration**

```{r Sulphates and More}

ggplot(aes(x = free.sulfur.dioxide, y=total.sulfur.dioxide), data = wqr) + 
  geom_point(alpha=1/2)+geom_smooth()+ylim(0,150)+xlim(0,60)

ggplot(aes(x = free.sulfur.dioxide, y=total.sulfur.dioxide+sulphates), 
       data = wqr) + geom_point(alpha=1/2)+geom_smooth()+ylim(0,150)+xlim(0,60)

#grid.arrange(sul1,sul2,ncol=2)

corFT <-with(wqr, cor.test(free.sulfur.dioxide,total.sulfur.dioxide))
corFS <- with(wqr, cor.test(free.sulfur.dioxide,sulphates))
corFTS <- with(wqr, cor.test(free.sulfur.dioxide,total.sulfur.dioxide + 
                               sulphates))
corFT$estimate
corFS$estimate
corFTS$estimate
               
modFTS <- lm(free.sulfur.dioxide ~ 
               total.sulfur.dioxide+sulphates+residual.sugar, data = wqr)
mtable(modFTS)
modFTS$coefficients

wqr$sulfcomb <- 0.2119922*wqr$total.sulfur.dioxide + 
  1.4209109*wqr$sulphates+5.0889418
corFTSmod<- with(wqr, cor.test(free.sulfur.dioxide,sulfcomb))
corFTSmod
```

Surprisingly, linear combination between *'sulphates'* and *'total.sulfur.dioxide'* did not improve correlation with *'free.sulfur.dioxide'*. Another dead-end where hypothesis and testing yielded a negative result. Ready to move on at this point.


**Bonus Obvervations 'X'**

A tip from a friend lead me to the following investigation. Is there any bias in the observation number and the quality of the wine? Perhaps the experts were tired and observations with larger X could indicate scoring later in the process assuming a preservation of the order.
```{r Observation}
#Create a factored version of the observations
wqr$x.cut<- cut(wqr$X, breaks = c(0,200,400,600,800,1000,1200,1400,1600))

ggplot(aes(y = X, x = factor(quality)), data = wqr) +
  geom_point(size = 3, alpha = 0.5, position = 'jitter') +
  geom_boxplot(alpha=0.25, fill = 'green')+
  geom_hline(yintercept = 800, linetype = 2, color = 'red') +
  ggtitle('Observations Vs Quality') +
  labs(y='Observation Number "X"', x='Quality')
  
```

Looking at the plot, it seems that aside from the exception of wines with *quality* 3, all other scoring seemed evenly distributed in terms of observations. The median tended towards the 800 mark as expected in a 1599 observation data set. One can be reasonably assured that at least there was no bias introduced through the ordering of the observations if the original order was preserved. 


**Principal Component Analysis**
```{r PCA}
#subset out the feature set for pca analysis

log.wqr <- log(wqr[, 2:12])
partial.wqr <- wqr[, 2:12]
redwine.quality <- wqr[, 14]
 
# apply PCA - scale. = TRUE is highly 
# advisable, but default is FALSE. 
redwine.pca <- prcomp(partial.wqr,
                 center = TRUE,
                 scale. = TRUE)

summary(redwine.pca)
print(redwine.pca)
```

Throughout this analysis, it always seemed that I might possibly be missing some hidden connections between the variables. Then I remembered that principal component analysis is one powerful way to reveal hidden structure within the data-set. It can help identify how variables work together, and maybe reduce the dimensionality of the data. Principal component 1 (PC1) contains only 28% of the total variance within the 11 possibly correlated 'independent' variables. And as for dimensionality reduction, PC1-PC9 contains > ~98% of the variance providing a dimensionality reduction of 2 using a cutoff of 95% variance captured.

Looking at the breakdown of first two principal components: PC1 is a principal component made up of the variables related to acidity and pH and PC2 is primarily related to sulfur dioxide. PC2 is interesting because this actually in keeping with the findings in the sulphate/sulfur dioxide exploration I performed above. Sulfur dioxide both free and total are major components of PC2 but sulphate contributes almost no variance to PC2. PC3 is the one that has some significant contribution from alcohol. 

The results of the principal component analysis reveals that there's not a super dominant component, this is also in keeping with the linear regression where it took a majority of the variables to get the best scores. I think I will use PCA much earlier in my exploratory data analysis in the future.                                                                                                                                                                                                                              


```{r PCA2}
#Preparing a dataframe to plot principal component Variances
pcaSum <- summary(redwine.pca)
varProp <- pcaSum$importance
tvarProp <- t(varProp)
tvarProp <- cbind(Row.Names = rownames(tvarProp), tvarProp)
tvarProp = as.data.frame(tvarProp)
names(tvarProp)[3]<-'Proportion.of.Variance'
names(tvarProp)[4]<-'Cumulative.Variance'
tvarProp$Cumulative.Variance = 
  as.numeric(as.character(tvarProp$Cumulative.Variance))
tvarProp$Proportion.of.Variance = 
  as.numeric(as.character(tvarProp$Proportion.of.Variance))

tvarProp <- transform(tvarProp, Row.Names = factor(Row.Names, 
    levels = c("PC1", "PC2", "PC3", "PC4", "PC5", "PC6", "PC7",
               "PC8","PC9","PC10","PC11"), ordered = TRUE))

#graphical representation of variance proportions of the principal components
ggplot(aes(x = Row.Names, y = Proportion.of.Variance, fill = Row.Names),
       data = tvarProp) + 
  geom_bar(stat='identity') +
  labs(x='Principal Components', y='Proportion of Variance') + 
  ggtitle('Principal Components Variance Proportions')


#prep pca analysis results for plotting
pcaRotation <- redwine.pca$rotation
pcaRotation = as.data.frame(pcaRotation)
rotNames <- cbind(Row.Names = rownames(pcaRotation), pcaRotation)
tidyPCA <-  tidyr::gather(rotNames, 'PC','Weights', 2:12)
tidyPCA$PCs = tidyPCA$variable
tidyPCAp2 <- 
  rbind(subset(tidyPCA, variable == 'PC1'), subset(tidyPCA, variable == 'PC2'), 
        subset(tidyPCA, variable == 'PC3'), subset(tidyPCA, variable == 'PC4'))

#graphical representation of the principal Component weights
ggplot(aes(x = Row.Names, y = value, color = PCs, group = PCs), 
             data = tidyPCA) + 
  geom_point() + geom_line(size = 1, alpha = .7, linetype = 2) + 
  labs(x='Features', y='PC Weights') + 
  ggtitle('Principal Component Weights Breakdown by PCs') +
  theme(axis.text.x = element_text(angle = 20, hjust = 1), 
        legend.text = element_text(size = 8))

pcaSum <- summary(redwine.pca)


```

The graphical representation of the results of the principal component analysis can allow for a quick overview of the top contributors to individual principal components as well as the proportions of and cumulative variance. The dotted lines for the *PC Weights Breakdown* graph makes it easier to trace the line. 

```{r}
#graphical representation of the principal Component weights
ggplot(aes(x = Row.Names, y = value, fill = PCs, group = PCs), 
             data = tidyPCA) + 
  geom_bar(stat='identity', position='dodge') + 
  labs(x='Features', y='PC Weights') + 
  ggtitle('Principal Component Weights Breakdown by PCs') +
  theme(axis.text.x = element_text(angle = 20, hjust = 1), 
        legend.text = element_text(size = 8))

ggplot(aes(x = Row.Names, y = value, fill = PCs, group = PCs), 
             data = tidyPCAp2) + 
  geom_bar(stat='identity', position='dodge') + 
  labs(x='Features', y='PC Weights') + 
  ggtitle('Principal Component Weights Breakdown by PCs') +
  theme(axis.text.x = element_text(angle = 20, hjust = 1), 
        legend.text = element_text(size = 8))
```

Line graphs tend to suggest a trending which is not the case with PCA weights, the bar graph is perhaps a more apt graph. I chose the line graph previously because I felt it show the information with more clarity, but perhaps displaying 12 PCs together is too cluttered and not altogether useful. A set of the most important PCs with their weight distributions are probably more useful.

###4: Final Plots and Summary

After the investigation of the data set and numerous dead ends, the two variables have seem to have the strongest affect on red wine quality are Volatile Acidity and Alcohol.

####Plot One
```{r Plot One}
ggplot(aes(y = sqrt(volatile.acidity), x = factor(quality) ), data = wqr) +
  geom_point(aes(color=cAcid.cut), position = 'jitter', size = 4, alpha = .75) +
  stat_summary(fun.y = mean, geom = 'point',size = 6, shape = 7,color = 'red') +
  geom_boxplot(alpha = 0.25)+
  scale_color_brewer(type = 'seq', palette = 'RdPu',
    guide = guide_legend(title = 'Citric Acid (g/dm^3)', reverse = F,
    override.aes = list(alpha = 1, size = 1)))+
  ggtitle('Red Wine Quality vs Volatile Acidity Colored by Citric Acid')+
  labs(x='Quality (1-10)', y='Volatile Acidity (acetic acid - g/dm^3)')

```
------ 
*Volatile Acidity* seems to be one of the primary contributors to quality, the box plot aids to reveal the linear nature of this input variable to the dependent. An added observation/visualization is that *Citric Acid* is negatively correlated with *Volatile Acidity*.

####Plot Two
*Principal Component Analysis*
```{r Plot Two, fig.height = 12}


p1 <- ggplot(aes(x = Row.Names, y = Proportion.of.Variance, 
                 fill = Row.Names, group = Proportion.of.Variance), 
             data = tvarProp) + geom_bar(stat='identity') +
  labs(x='Principal Components', y='Proportion of Variance') + 
  ggtitle('PCA Variance Proportions') +
  theme(axis.text.x = element_text(angle = 25, hjust = 1)) +
  scale_y_continuous(limits = c(0, 0.3)) + 
  scale_fill_brewer(type = 'seq', palette = 'Spectral',
    guide = FALSE)

p2 <- ggplot(aes(x = Row.Names, y = Cumulative.Variance), 
             data = tvarProp) + geom_bar(stat='identity') +
  labs(x='Principal Components', y = 'Cumulative Variance') + 
  ggtitle('PCA Cumulative Variance') +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))


p3 <- ggplot(aes(x = Row.Names, y = value, fill = PCs, group = PCs), 
             data = tidyPCAp2) + 
  #geom_point() + geom_line(size = 1, alpha = .7, linetype = 2) + 
  geom_bar(stat='identity', position = 'dodge') +
  labs(x='Features', y='PC Weights') + 
  ggtitle('Principal Components Weight Breakdown of First Four PCs') +
  theme(axis.text.x = element_text(angle = 20, hjust = 1), 
        legend.text = element_text(size = 8)) +
  scale_fill_manual(values=c("#9e0142", "#d53e4f", "#f46d43","#fdae61"),
                     guide = guide_legend(title = 'Citric Acid', reverse = F,
    override.aes = list(alpha = 1, size = 3)))

grid.arrange(arrangeGrob(p1,p2, ncol=2), p3, ncol = 1, heights = c(1,1.5))
```

The graphical representation of the principal component analysis (PCA) helps to quickly visualize the distribution of the total variance in the data set. The bar graph showing the variance proportions for each PC and the color coordinated line graph showing the relative contributions of each variable to the respective PCs can give a quick way to find hidden relationships. The main relationships are acidity, sulfur dioxide, and alcohol corresponding to PC1, PC2 and PC3. It takes nine principal components to capture more than 95% of the total variance.


####Plot Three
```{r Plot Three}
ggplot(aes(y = sqrt(alcohol), x = factor(quality)), data = wqr)+
  geom_point(aes(color = x.cut ),position = 'jitter', size = 3, alpha = .5)+
  stat_summary(fun.y = mean, geom = 'point',size = 6, shape = 7,color = 'red') +
  geom_boxplot(alpha = 0.25)+
  scale_color_brewer(type = 'seq', palette = 'Greens',
    guide = guide_legend(title = 'Observation Number "X"', reverse = F,
    override.aes = list(alpha = 1, size = 3)))+
  ggtitle('Red Wine Quality vs Alcohol with Colored Means')+
  labs(x='Quality (1-10)', y='Alcohol (% by volume)')

```
------ 
The final low hanging insight is that *alcohol* is linearly correlated with quality and should be a significant contributor if one was to build a predictive model around the given variables and red wine quality. It seems that at least for red wine, higher alcohol content means better quality score! The chart is also colored by the observation number *'X'* to demonstrate that there is no bias in the inherent ordering of the observations (as in observations numbered 1-200 do not have a tendency to have higher quality ratings than observations 1400-1600).



###5: Reflections

Exploring the Red Wine data-set has been somewhat frustrating. After coming from the lesson and the Diamonds and Facebook data-set where intuition paid off at least two times out of three, the red wine's data refused to give up anything that was not somewhat obvious. Working with a qualitative/unbalanced sampling of the output/dependent variable was somewhat challenging as well. I keep thinking that discarding the bottom 2 and the top levels of the quality factor would vastly improve correlations at the risk of making the modeling trivial and losing all meaning to the data. The variables that affect Red Wine Quality most are *'alcohol'*, *'volatile.acidity'* and *'sulphates'*. *'citric.acid'* though correlated reasonably high with quality has high correlation with *'volatile.acidity'* and loses much of its impact on the linear modeling due to the relationship. For the future, perhaps a non-linear model could be implemented and tested. Certain variables that I had hoped would show better correlation through informed factoring did not perform as expected (*'free.sulfur.dioxide'*). Instead *'total.sulfur.dioxide'* gave better correlation after factoring. The exploration of the red wine data-set has surprisingly taken more time than I expected. In the future I also think I will use Principal Component Analysis earlier in the exploration to quickly see if there are hidden or even more obvious relationships between variables. R is truly a powerful data analysis tool.


